Hilbert space methods to approximate Gaussian processes using Stan

### 2019-02-25 Paper/manuscript/manuscript_05.tex
improve introduction

Gaussian processes (GPs) are natural and flexible non-parametric models to define probability distributions over multi-dimensional functions \citep{rasmussen2006gaussian,neal1997monte}. This enables to apply probabilistic inference directly to functions. The defining feature of a GP model is that the joint distribution of function values is a multivariate normal distribution which is characterized by a mean function and a covariance function. The covariance function encodes our prior assumptions about the function, such as continuity, smoothness, periodicity and scale properties of the GP, and defines a correlation structure which characterize the correlations between function values at different inputs. A GP model with a covariance function allows for non-linear effects on the function values and can handle interactions between input variables implicitly. Several types of covariance functions with different behaviors can be used, from stationary (e.g. exponentiated quadratic, Matern, periodic, quasyperiodic...) to non-stationary (e.g. dot product, neuronal network,...). Combinations, sum and product, of different covariance function can also be used in order to obtain more modelling flexibility on real complex problems. GP models are often referred to as non-parametric because the number  of parameters is not fixed, but rather spanned by the number of data points, which allows to adapt model complexity to the data. GP models are powerful non-parametric prior models for regression and classification \citep{rasmussen2006gaussian}, and they can also be used as non-linear components in additive models. GP models is of broad interest across machine learning and statistics, with applications in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, Bayesian optimization and probabilistic numerics, \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}, and many others.

One of the main limitations of GPs is that computational demands and memory requirements scale as $O(n^3)$ and $O(n^2)$ respectively, effectively limiting a direct implementation to a number of training samples $n$ at most a few thousand cases.   This problem becomes especially important when using sampling methods for full Bayesian inference, when in each sampling step we need to invert the Gram matrix, usually through Cholesky factorization.

In this paper we want to make full Bayesian inference over Gaussian process models accessible to a broader audience by making them fast and easy to apply. We implement novel method recently  developed by \cite{solin2018hilbert} for reduced-rank approximations of GP models. This method decomposes the GP kernel into basis functions which represent the GP as a linear model and makes it faster to fit. The set of Laplace eigenfunctions considered in this paper forms such a basis and we can compute them analytically. Furthermore, they are independent of the choice of kernel (including the hyperparameters), which means that we do not need to autodiff through the kernel and the Cholesky. In a fully Bayesian inference framework using sampling methods, the proposed approximate GP model has a computational complexity of $O(nm+m)$ in every step of the HMC method. The parameter posterior distribution in this approximate GP model is $m$-dimensional ($m<<n$) which helps the use of GP priors as latent functions, since it is known that there exist a correlation between the latent values and the hyperparameters which makes inference more difficult especially when the dimension of latent values is high.

There already exist several approximate methods that alleviate these computational demands. Sparse GPs, which are approximate methods to fit GP models, are based on building a low-rank approximation to the covariance matrix of the complete data based around 'inducing variables'. They do reduce the dimension of the posterior, reducing the storage requirements to $O(nm)$ and complexity to $O(nm^2)$, where $m < n$. However it still requires Cholesky factorizations. Furthermore convergence to the full GP model is not straight forward, and reducing the amount of data while maintaining the posterior densities is not a simple procedure. Some of these methods have been reviewed in \cite{rasmussen2006gaussian}, and \cite{quinonero2005unifying} provide a unifying view of these methods based on approximate generative methods. More recently Bui et. al (2017) revisit the inducing points-based sparse approximation methods, in which all the necessary approximation is performed at inference time, rather than at the modelling time. \cite{wilson2015kernel} introduce a new unifying framework for inducing point methods, called structured kernel interpolation (SKI).

Other global approach make use of the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. It is well-known (see, e.g., \cite{wahba1990spline}) that spline smoothing can be seen as Gaussian process regression with a specific choice of covariance function, but they do not generally have correspondence with GP models, and they do not have the flexibility of choosing different covariance functions as model structure. The Sparse Spectrum GP is based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving,gal2015improving}. Recently \citep{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat´ern class of kernels. Random Fourier Features \citep{rahimi2008random,rahimi2009weighted} is a method for approximating kernels.

The approach we propose in this paper is based on the novel method developed by \cite{solin2018hilbert} for reduced-rank approximations of GP models. This method is based on interpreting the covariance function as the kernel of a pseudo-differential operator and approximating it using Hilbert space methods. This results in a reduced-rank approximation for the covariance function.

While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank.
Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.

There are several GP specific software, which have efficient
computation for restricted set of GP based models allowing scaling to
bigger data sets. In this paper we do not focus on the fastest
possible inference for a specific restricted models, but instead we
are interested in how GPs can be easily used as modular components in
probabilistic programming frameworks.

As said above the proposed method was already developed by \cite{solin2018hilbert} where they fully develop, describe and generalize the methodology. Though, they do not put much effort in describing and analyzing the relation among the key factors of the box size (or boundary condition), the number of basis functions, and the smoothness or roughness of the function. The performance and accuracy of the method are directly related with these factors. The time of computation is mainly dependent on the number of basis functions. Our main contributions to this recently developed methodology for low-rank GP model by \cite{solin2018hilbert} goes around these aspects. We investigate the relations going on among these factors, the number of basis functions, the box size, and the lengthsccale of the functions.  We make recomendations for the values of these factors based on the recognized relations among them. We provide useful graphs of these relations that will help the users to improve performance and save time of computation. We also diagnose if the chosen values for the number of basis functions and the box size are adequate to fit to the actual data.

The rest of the paper is structered as follows. Section 2 describe  the proposed model based on the Hilbert space methods for low-rank approximation to GP. In Section 3 we analyze the accuracy of the method and its dependency on existing relations among the number of basis functions, the box size factor and the roughness or smoothness of the process.

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive, because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in eq. (\ref{eigenproblem}) are sinusoidal functions,
%
\begin{eqnarray}
\lambda_j&=&\left(\frac{j\pi}{2L}\right)^2, \nonumber \\
\phi_j(x)&=&\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right). \nonumber
\end{eqnarray}

@@

The approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, e.g. in a 1-$D$ input space $x\in \Omega=[-L,L]$:
%
\begin{equation}\label{approxcov}
k(\tau) \approx \sum_{j}S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

>>>>

Gaussian processes (GPs) are natural (Paul: what does "natural" imply and as compared to what?) and flexible non-parametric models to define probability distributions over multi-dimensional functions \citep{rasmussen2006gaussian,neal1997monte}. This allows applying probabilistic inference to estimate functional relationships between one or more covariates and a single response variable. The defining feature of GPs is that the function values are assumed to jointly follow a multivariate normal distribution, which is characterized primarily by a covariance function, sometimes also referred to as covariance kernel. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Several types of covariance functions with varying properties can be used, and they may even be combined for further increased flexibility. GPs are often referred to as non-parametric models because the number of parameters is not fixed, but rather increases with the number of data points, which allows to adapt model complexity to the data. As such, the term 'non-parametric' does not imply having no parameters but actually having a lot of them. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

One of the main limitations of exact GPs is that computational demands and memory requirements scale as $O(n^3)$ and $O(n^2)$, respectively, with the number $n$ of observations in the data used to fit the model. This effectively limits their application to rather small data sets of a few thousand observations at most. This problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we need to invert the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of 'inducing variables' (Paul: if you introduce this term, you may need to explain it. Alternatively, can we just get rid of it?). They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. More recent developments in the context of sparse GPs include points-based sparse approximation methods \citep{wilson2015kernel}, which have been further developed by Bui et. al (2017) (Paul: cite this study correctly in latex) to perform approximations at inference time rather than at modeling time.

(Paul: This whole paragraph is unclear to me. Are all of these approaches approximations via splines, or are the splines just special cases of a more general method? In any case, I think the paragraph needs substantial revision). Other global approach make use of the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. It is well-known (see, e.g., \cite{wahba1990spline}) that spline smoothing can be seen as Gaussian process regression with a specific choice of covariance function, but they do not generally have correspondence with GP models, and they do not have the flexibility of choosing different covariance functions as model structure. The Sparse Spectrum GP is based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving,gal2015improving}. Recently \citep{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat´ern class of kernels. Random Fourier Features \citep{rahimi2008random,rahimi2009weighted} is a method for approximating kernels.

Yet another approach to approximate GPs -- on which we will focus in the present paper -- was proposed by \cite{solin2018hilbert} who suggested to use reduced-rank approximations of the covariance kernel. This method decomposes the kernel into basis functions which represent the GP as a linear model and thus makes its estimation considerably faster. In principal, we are free in the choice of basis functions, but the Laplace eigenfunctions were found to be particularily appealing \citep{solin2018hilbert}. Not only can they be computed analytically, but they are also independent of the particular choice of the covariance kernel including the hyperparameters. The latter comes with several advantages when performing inference via algorithms that require differentiation of the posterior density (Paul: cite something here). In a fully Bayesian framework based on sampling methods, the proposed approximate GP model has a computational complexity of just $O(nm+m)$, with $m<<n$ in every sampling step. The posterior distribution of the reduced-rank GP approximation is of dimension $m$ and thus considerably smaller than in the case of exact GPs. This not only lowers the memory requirements but also reduces the correlations between latent values and the corresponding hyperparameters of the GP, which can help to further improve sampling efficiency.

While \cite{solin2018hilbert} fully developed the mathematical theory behind reduced-rank approximations of GPs, they did not put much effort in analyzing the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space (Paul: I replace 'box size' because, at that point, readers will likely have no idea what that means), or properties of the true functional relationship between covariates and response variable. However, for practical implementations of approximate GPs, the latter factors are critical and thus require further investigation. In the present paper, we will fill this gap by providing practical recommendations for the choice of these factors based on the recognized relations among them. What is more, we provide intuitive visualizations of these relations that will help users to improve performance and save computation time, while at the same time maintaining close approximations of exact GPs.

Although there are several GP specific software packages available to date (Paul: cite those), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for a specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan (cite). To this end, we want to make full Bayesian inference over Gaussian process models accessible to a broader audience by making them easy, fast and save to apply.

The remainder of the paper is structered as follows. In Section 2, we introduce GPs and their reduced rank approximations proposed by \cite{solin2018hilbert}. In Section 3, we analyze the accuracy of these approximations under several conditions using analytical and numerical methods. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Section 4. We end with a discussion in Section 5.

(Paul: The content and the purpose of the following sentences is not fully clear to me. Do we need them in the Introduction or can they be removed, or put into the next section?) While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive, because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in eq. (\ref{eigenproblem}) are sinusoidal functions,
%
\begin{eqnarray}
\lambda_j&=&\left(\frac{j\pi}{2L}\right)^2, \\
\phi_j(x)&=&\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right).
\end{eqnarray}

@@

The approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, e.g. in a 1-$D$ input space $x\in \Omega=[-L,L]$:
%
\begin{equation}\label{diffcov}
k(\tau) \approx \sum_{j}S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}
